---
title: "Assignment 10: Data Scraping"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r, message = FALSE}
#1
#install.packages("rvest")
library(rvest)
library(tidyverse)

getwd()

my_theme <- theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(colour = "#440000"),
    plot.title = element_text(colour = "blue",
                              hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title = element_text(colour = "#4169e1")
  )
theme_set(my_theme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
durham_lwsp.web <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "36.1000".

```{r scrape.the.data}
#3
water.system.name <- durham_lwsp.web %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
water.system.name

PWSID <- durham_lwsp.web %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
PWSID

ownership <- durham_lwsp.web %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
ownership

max.withdrawals.mgd <- durham_lwsp.web %>%
  html_nodes("th~ td+ td") %>%
  html_text()
max.withdrawals.mgd
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the average daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4
months <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
monthly.withdrawal.df <- data.frame(water.system.name, 
                                    PWSID,
                                    ownership,
                                    months,
                                    max.withdrawals.mgd) %>%
  rename(withdrawal = max.withdrawals.mgd) %>%
  mutate(withdrawal = as.numeric(withdrawal),
         months = lubridate::month(months))

#5
axis.label <- as.vector(sort(lubridate::month(months, label = TRUE)))

ggplot(data = monthly.withdrawal.df,
       mapping = aes(x = months,
                     y = withdrawal)) +
  geom_point(colour = 'red') +
  geom_line(colour = 'black') +
  scale_x_discrete(limits = axis.label) +
  labs(x = "Months",
       y = "Withdrawal (MGD)",
       title = "Average Daily Withdrawals Across 2022")
                     

```
\newpage
6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.

#https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022
#url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", url.pwsid, "&year=", url.year)
#url

scraperboi <- function(url.pwsid, url.year) #This function takes the URL from ncdeq(), scrapes the website and returns a dataframe with name of city, PWSID, ownership, and the withdrawal for each month of the year.
{
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", url.pwsid, "&year=", url.year)
  
  water_supply <- read_html(url)

  #scraping the webpage for the data we need
  water.system <- water_supply %>%
    html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
    html_text()

  PWSID <- water_supply %>%
    html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
    html_text()

  ownership <- water_supply %>%
    html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
    html_text()

  max.withdrawals <- water_supply %>%
    html_nodes("th~ td+ td") %>%
    html_text()

  scaped.months <- c(01, 05, 09, 02, 06, 10, 03, 07, 11, 04, 08, 12)
  
  monthly.withdrawal <- data.frame(water.system,
                                   PWSID,
                                   ownership,
                                   url.year,
                                   scaped.months,
                                   max.withdrawals) %>%
    rename(withdrawal = max.withdrawals,
           months = scaped.months,
           year = url.year) %>%
    mutate(withdrawal = as.numeric(withdrawal),
           months = lubridate::month(months, label = FALSE))
  return(monthly.withdrawal)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
#these two variables serve as input arguments for ncdeq() which is called by scraperboi() 
url.pwsid <- "03-32-010"
url.year <- "2015"

durham.monthly.withdrawal <- scraperboi(url.pwsid, url.year)

#plotting the results

ggplot(data = durham.monthly.withdrawal,
       mapping = aes(x = months,
                     y = withdrawal)) +
  geom_point(colour = 'red') +
  geom_line(colour = 'black') +
  scale_x_discrete(limits = axis.label) +
  labs(x = "Months",
       y = "Withdrawal (MGD)",
       title = "Average Daily Withdrawals in Durham Across 2015")

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
asheville.monthly.withdrawal <- scraperboi('01-11-010', 2015)

#plotting the results

ggplot(data = asheville.monthly.withdrawal,
       mapping = aes(x = months,
                     y = withdrawal)) +
  geom_point(colour = 'blue') +
  geom_line(colour = 'black') +
  scale_x_discrete(limits = axis.label) +
  labs(x = "Months",
       y = "Withdrawal (MGD)",
       title = "Average Daily Withdrawals in Asheville Across 2015")

legend.colours <- c("durham" = "lightblue", "asheville" = "orange")
ggplot() +
  geom_line(data = durham.monthly.withdrawal,
            mapping = aes(x = months,
                          y = withdrawal,
                          colour = "durham"),
            size = 1) +
  geom_line(data = asheville.monthly.withdrawal,
            mapping = aes(x = months,
                          y = withdrawal,
                          colour = 'asheville'),
            size = 1) +
  scale_x_discrete(limits = axis.label) +
  scale_color_manual(values = legend.colours) +
  labs(x = "Months of 2015",
       y = "Withdrawal (MGD)",
       title = "Comparison of average monthly withdrawal of Asheville and Durham",
       colour = "Legend")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
years <- seq.int(from = 2010, to = 2021, by = 1)
location.pwsid <- rep("01-11-010", 12)

asheville.withdrawals.2010.2021 <- map2(location.pwsid,
                                        years,
                                        scraperboi) %>%
  bind_rows() %>%
  arrange(year, months) %>%
  mutate(year = as.factor(year))

ggplot(data = asheville.withdrawals.2010.2021,
       mapping = aes(x = months,
                     y = withdrawal,
                     colour = year)) +
  geom_point() +
  geom_smooth(method = "loess") +
  #geom_line() +
  scale_x_discrete(limits = axis.label) 

```

> Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
>We see that the water usage has a consistent seasonal component across the years in our sample size. Consumption is highest in July, decreases through autumn and winter to its lowest point in March.
