---
title: 'Assignment 2: Coding Basics'
author: "Nagarajan Vaidya Subramanian"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. Using the seq function from the base package to generate a sequence of numbers from 1 to 100 in increments of 4
num_seq <- seq(1, 100, by=4)
#2. calculating the mean and median of this sequence
mean.num_seq <- mean(num_seq)
mean.num_seq
median.num_seq <- median(num_seq)
median.num_seq

#3. Using an if condition to check whether mean is greater than the median
mean.num_seq > median.num_seq
#if (mean.num_seq > median.num_seq) {
#  print("Mean of the sequence is greater than the median")} else {
#  print("Mean of the sequence is not greater than the median")}
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
student_names <- c("a", "b", "c", "d") #vector containing 4 names of students
test_scores <- c("54", "95", "12", "70") #vector containing test scores (random numbers)
test_grades <- test_scores>=50 #vector containing grades of the 4 students based on their test scores
marksheet.df <- data.frame(student_names, test_scores, test_grades) #creating a data frame from the three vectors we created
colnames(marksheet.df) <- c("names", "scores", "grades") #changing the column names to something meaningfull but easier to write than the vector names themselves
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A matrix is a simple collection of numbers arranged across rows and columns. Whereas a data frame is a more sophisticated data structure that can store a mix of data types along with column/row names.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
passingGrade <- 50 #variable to store the passing score of this test
passingTest <- function(vect, pass, print=TRUE) {
  for (i in 1:length(vect)) {
    if (vect[i]>=pass) {
      print(paste(vect[i], "is a passing grade"))
    } else {
      print(paste(vect[i], "is not passing grade"))
    }
  }
}
passingTest(test_scores, passingGrade)
```

```{r}
passingTest2 <- function(vect, pass, print=TRUE) {
  for (i in 1:length(vect)) {
    ifelse (vect[i]>=pass, print(paste(vect[i], "is a passing grade")), print(paste(vect[i], "is not passing grade")))
  }
}
passingTest2(test_scores, passingGrade)
```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: I tried both, and got the same result (as we can see above)
